---
title: "Signal Detection Theory"
format: 
  revealjs:
    smaller: true
    incremental: true
editor: visual
---

## The Yes-No task

. . .

| Stimulus | Response |            Correct?            |
|:--------:|:--------:|:------------------------------:|
|    S     |   Yes    |        Correct: **Hit**        |
|    N     |    No    | Correct: **Correct Rejection** |
|    S     |    No    |      Incorrect: **Miss**       |
|    S     |    No    |      Incorrect: **Miss**       |
|    N     |   Yes    |   Incorrect: **False Alarm**   |
|    S     |   Yes    |        Correct: **Hit**        |

|              |     |     |
|:------------:|:---:|:---:|
|              |  N  |  S  |
| Yes, it is S |     |     |
| No, it is N  |     |     |

## Binomial distribution: probability

. . .

A randow variable $Y$ is distributed binomial if the pdf is

$$f(y;n, p) = \binom{n}{y} p^y \, (1-p)^{n-y}$$

-   We have an experiment that has two possible outcomes (head or tail).
-   We have a probability $p$ that one of the events happens. Let's say 0.8 for heads.
-   We repeat the experiment $n$ times. Let's say 50.

. . .

$f(y;n, p)$ corresponds to the probability of obtaining $y$ heads.

. . .

For example, the probality of obtaining 25 heads is:

```{r}
#| echo: true
dbinom(25, size = 50, prob = 0.8)
```

Very small

```{r}
#| echo: true
dbinom(45, size = 50, prob = 0.8)
```

## Binomial distribution: statistics

We have the data: tossing a coin 200 times and getting 120 heads.

. . .

Which is the most likely value of the parameter $p$?

. . .

We need to compute a statistic from the data.

. . .

$$\hat{p} = ?$$

. . .

$$\hat{p} = \frac{n_h}{n} = \frac{120}{200} = 0.6$$
